---
title: "lists"
author: "JP"
date: "September 12, 2017"
output: html_document
---
From [R4DS 20.5](http://r4ds.had.co.nz/vectors.html#lists)
```{r}
library(tidyverse)
```

```{r generate example list}
a <- list(a = 1:3, b = "a string", c = pi, d = list(-1, -5))
```

## 1) `a[1]` uses `[` *to extract a sub-list, which will always be of type "list".
```{r extract a sub-list}
a[1]
a[1] %>% str

str(a[4])

a[4]
```

The output says
"List of 1
 $ d:List of 2"
I think this refers to `a` being a list of one vector, and `list component` 4, which is `named element` d, is a list of 2 vectors.

## 2) `a[[1]]` uses `[[` *to extract a single component from a list*. It removes a level of hierarchy from the list. In this case producing an atomic vector.
```{r extract a single component}
a[[1]]

# same below
a[["a"]]
```

## 3) `a$a` uses `$` as shorthand for *extract named element of list*. It works similarly to `[[` except that you don’t need to use quotes.
```{r extract named element}
a$a

a[["a"]] # same as a[[1]]
```