---
title: "spread and gather"
author: "JP"
date: "Aug 10, 2017"
output: github_document
---

```{r}
library(tidyverse)
```

## gather and spread examples
```{r}
# create some data to spread and gather
mydata <- tribble(
  ~group_a, ~group_b, ~group_c, ~group_d, 
  21,       14,       10,       23,
  24,       20,       12,       35,
  34,       45,       34,       14
)

mydata

# Tidy up the data
mydata_gathered <- 
  mydata %>% 
  mutate(row = row_number()) %>% # rownumber is key for later spread b/c you
  # need a variable that uniquely identifyies observations
  gather(
    group_a:group_d, # old variables to gather
    key = group,     # new variable whose values will be old variable names
    value = score    # new variable whose values will be old variable values
  ) 

mydata_gathered
```

```{r}
# demo a spread
mydata_spread <- 
  mydata_gathered %>% 
  spread(key = group, value = score)

mydata_spread
```

"row number `spread()`" soln from [link](https://stackoverflow.com/questions/25960394/spread-with-data-frame-tibble-with-duplicate-identifiers).

`convert = TRUE` will produce variables of mixed types for `spread()` and detect types for `gather()`.

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks %>% 
  spread(
    key = year, 
    value = return) 

stocks %>% 
  spread(key = year, value = return) %>% 
  gather(key = year, value = return, `2015`:`2016`, convert = TRUE) %>% 
  select(year, half, return)
```

From R4DS 12.3.3 
Why are `gather()` and `spread()` not perfectly symmetrical?
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)
```

Gathered variables default to character class, and do not return to original class when re-spread.  

Both spread() and gather() have a convert argument.
```{r}
stocks %>% 
  spread(year, return, convert = TRUE) 

stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`, convert = TRUE) 
```

Code below fails because variables with numeric names require back ticks, e.g., `1999`.
```{r message=FALSE, warning=FALSE, eval=FALSE}
table4a %>% 
  gather(1999, 2000, key = "year", value = "cases")
```


The `spread()` fails in the tibble below because of duplicate identifiers in the `names` column. This is because  Phillip Woods has two ages listed.  
- Adding row numbers will fix the problem, but create a long and sparse dataset.  
- Resolving the age double entry for Woods is a better soln. 
```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
```

```{r eval=FALSE}
people %>% spread(key = key, value = value)
```

```{r}
people %>% 
  mutate(row_num = seq(1:nrow(.))) %>% 
  spread(key = key, value = value)

people %>% 
  mutate(
    value = ifelse(name == "Phillip Woods" & key == "age", 47.5, value)
  ) %>% 
  filter(!duplicated(.)) %>% 
  spread(key = key, value = value)
```

Tidy the simple tibble below. Do you need to spread or gather it? What are the variables?
```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)

preg %>% 
  gather(male, female, key = gender, value = number) %>% 
  select(gender, pregnant, number)
```

These data can be tidied via a gather and the new variables are gender, pregnant, and number. 
